---
title: "Formula Interface for ggplot2"
author: "Daniel Kaplan and Randall Pruim"
date: "February 25, 2016"
output:  rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Formula Interface for ggplot2}
---

```{r include = FALSE}
library(ggformula)
library(dplyr)
library(ggplot2)
library(mosaic)
theme_set(theme_light())
```


# Formula-driven graphics


One of the main materials from which statistical models are made is graphics. A data graphic is a representation of a data table which has particular resonance with human cognitive abilities and which builds on topics found in widely taught math curricula, e.g. cartesian coordinates.

There are many excellent graphics packages provided for R. The `statisticalModeling` package currently builds on one of them, `ggplot2`. The purpose of the graphics functions in `statisticalModeling` is to provide a clear, concise, and consistent interface to the underlying graphics package.

A simple example:
```{r}
library(ggformula)
gf_point(mpg ~ hp, data = mtcars)
```

Each of the marks in the plot is a *glyph*. Every glyph has graphical *attributes*. 
In the above plot, the attributes are x- and y-position. 
Other possible graphical attributes for a point glyph are color, shape, size, 
stroke, fill, and alpha (transparency).


Experts, of course, will want to use the underlying `ggplot2` package directly
in order to maximize expressiveness and minimize the "bureaucracy" induced by
intervening layers of software. The `ggformula` graphics are suited
for beginners because they reduce the number of elements used in notation.

There are three aspects to describing graphics using `ggformula`:

1. The "kind of graphic" to be created, e.g. scatter plots, density plots, bar plots, maps, etc.
2. The role that each variable plays in the graphic, e.g. x- or y-axis, color, shape, facet.
3. And, of course, the data table(s) on which the graphic is to be based.

The "kind of graphic" is specified by the name of the graphics function. All of
the `ggformula` data graphics functions have names starting with
`gf_`, which is intended to remind the user that they are formula-based
interfaces to `ggplot2`: `g` for `ggplot2` and `f` for "formula." Commonly used
functions are

- `gf_point()` for scatter plots
- `gf_line()` for line plots (connecting dots in a scatter plot)
- `gf_density()` or `gf_dens()` or `gf_histogram()` or `gf_freqpoly()` to display distributions
- `gf_boxplot()` for comparing distributions side-by-side
- `gf_counts()` for bar-graph style depictions of counts.
- `gf_bar()` for more general bar-graph style graphics

The function names generally match the corresponding operation in `ggplot2`,
although `gf_counts()` is a simplified special case, and `gg_dens()` is an 
alternative to `gg_density()` that displays the density plot slightly differently
than the default in `ggplot2`.

Each of the `gf_` functions can create the coordinate axes and fills it in one
operation. (In `ggplot2` nomenclature, `gf_` functions create a frame and add a
geom layer, all in one operation.)  This is what happens for the first
`gf_` function in a chain.  For subsequent `gf_` functions, new layers are added.

In the `gf_` functions, you specify the graphical attributes using a formula.
Each such specification takes the form `attribute:value` 
where `attribute` is one of `color`, `shape`, etc. and `value` is either a constant 
(e.g. `"red"` or `0.5` as appropriate) or a variable (e.g. `cyl`). Altenatively,
if the specification has the form `attribute::value`, then `value` may be
an expression that can be evaluated in the data (although often is is better
to create a new variable in the data and use `:`). For instance:
```{r}
gf_point(mpg ~ hp + color:cyl + size:carb + alpha:0.75, data = mtcars) %>%
  gf_labs(x = "Miles per gallon", y = "Horsepower", 
          title = "Horsepower and fuel efficiency")
```

```{r}
gf_point(mpg ~ hp + color:cyl, data = mtcars)
gf_point(mpg ~ hp + color:cyl + size:carb + alpha:0.75, data = mtcars)
gf_point(mpg ~ hp + color::factor(cyl) + size:carb + alpha:0.75, data = mtcars)
gf_point(mpg ~ hp + color:cylinders + size:carb + alpha:0.75, 
         data = mtcars %>% mutate(cylinders = factor(cyl)))
```

You add labels and facets using the standard `ggplot2` notation.  `xlab()` and `ylab()` are adding the axis labels in the above. (For faceting, see below.)

## Density and related plots

To illustrate, we'll use the density style `gf_density()`, but the same instructions apply to `gf_histogram()` and `gf_freqpolygon()`.  *Note that in the one-variable density graphics, the variable whose density is to be calculated goes to the right of the tilde, in the position reserved for the x-axis variable.*

```{r fig.show = "hold", out.width = "30%", warning=FALSE}
data(Runners, package = "statisticalModeling")
Runners <- Runners %>% filter( ! is.na(net))
gf_density( ~ net, data = Runners)
gf_density( ~ net + fill:sex + alpha:0.5, data = Runners)
gf_dens( ~ net + color:sex + alpha:0.5, data = Runners)    # no line on the bottom
gf_dens( ~ net + color:sex + alpha:0.5, data = Runners, adjust = 0.25)  # less smoothing
gf_dens( ~ net + color:sex + alpha:0.5, data = Runners, adjust = 4)     # more smoothing
```

When the `fill` or `color` or `group` aesthetics are mapped to a variable, the default behavior is to lay the group-wise densities on top of one another. Other behavior is also available by using `position` in the formula. Using the value `"stack"` causes the densities to be laid one on top of another, so that the overall height of the stack is the density across all groups. The value `"fill"` produces a conditional probability graphic.
```{r fig.show = "hold", warning=FALSE}
gf_density( ~ net + fill:sex + color:NA + position:"stack", data = Runners)
gf_density( ~ net + fill:sex + color:NA + position:"fill", data = Runners, adjust = 2)
```

Similar commands can be constructed with `gf_histogram()` and `gf_freqpoly()`,
but note that `color`, not `fill`, is the active aesthetic for frequency
polygons and `position:"fill"` doesn't work.

Box and whisker plots show the distribution of a quantitative variable as a
function of a categorical variable. The formula used in `gf_boxplot()` should
have the quantitative variable to the left of the tilde.

```{r fig.show = "hold", warning = FALSE}
gf_boxplot(net ~ sex + color:"red", data = Runners)
gf_boxplot(net ~ sex + color:start_position, data = Runners)
```

This plot may surprise you.

```{r}
gf_boxplot(net ~ year, data = Runners)
```
The warning message suggests that `gf_boxoplot()` is not really designed to be used
with quantiative data.  It suggests sets `group`, and sometimes, this is 
just what we want.
```{r}
gf_boxplot(net ~ year + group:year, data = Runners)
```

But often, is is better to convert a  discrete quantitative variable used for grouping into
a categorical variable. This can be done in several ways:
```{r}
Runners$the_year <- as.character(Runners$year)               # in base R
Runners <- Runners %>% mutate(the_year = as.character(year)) # in dplyr
gf_boxplot(net ~ the_year + color:sex, data = Runners)
# or do it all at once
gf_boxplot(net ~ as.character(year) + color:sex, data = Runners)
```

Two-dimensional plots of density also have both a left and right component to the formula.
```{r fig.show = "hold"}
gf_density_2d(net ~ age, data = Runners)
gf_hex(net ~ age, data = Runners)
```

## Paths and lines

The `ggplot` system offers two ways to connect points. `gf_line()` ignores the order of the points in the data draws the line going from left to right. `gf_path()` goes from point to point according to the order in the data. Both forms can use a `color` or `group` aesthetic as a flag to draw groupwise lines.

```{r fig.show = "hold", out.width = "30%"}
# create a categorical variable
mtcars <- mtcars %>% mutate(n_cylinders = as.character(cyl)) 
gf_line(mpg ~ hp, data = mtcars)
gf_path(mpg ~ hp, data = mtcars)
gf_line(mpg ~ hp + color:n_cylinders, data = mtcars)
```

The above are examples of *bad plots*. The viewer is unnecessarily distracted by the zigs and zags in the connecting lines. It would be better to use `gf_point()` here, but then you wouldn't see how `gf_line()` and `gf_path()` work!

## Multiple layers and chaining

Often it is useful to overlay multiple layers onto a single plot.  This can be done 
by chaining them with `%>%`, the "then" operator from `magrittr`.  The `data` 
argument can be omitted if the new layers uses the same data as the first layer 
in the chain. 

The following plot illustrates how histograms and frequency polygons are related.

```{r}
gf_histogram( ~ age, data = Runners, alpha = 0.2, fill = "navy") %>%
  gf_freqpoly( ~ age)
```

A 2-d density plot can be augmented with a scatterplot.

```{r}
gf_density_2d(net ~ age, data = Runners) %>%
  gf_point(net ~ age + alpha:0.03) 
```



## Faceting

The `ggplot` system allows you to make subplots --- called "facets" --- based on the values of one or two categorical variables. This is done by "adding" a `facet_grid()` directive. The directive uses a formula to specify which variable(s) are to be used in the grid.

```{r fig.show = "hold", warning=FALSE}
gf_density_2d(net ~ age, data = Runners) %>% gf_facet_grid( ~ sex)
gf_density_2d(net ~ age, data = Runners) %>% gf_facet_grid(start_position ~ sex)
```

## Global plot adjustments

There are a number of things we may want to do to the entire plot -- adjusting
labels, colors, fonts, etc.  `ggformula` provides wrappers to the `ggplot2` functions 
for this so that the chaining syntax can be used.

```{r}
gf_histogram( ~ age, data = Runners, alpha = 0.2, fill = "navy") %>%
  gf_freqpoly( ~ age) %>%
  gf_labs(x = "age (years)", title = "Age of runners") %>%
  gf_lims(x = c(20, 80)) %>%
  gf_theme(theme = theme_minimal)

gf_histogram( ~ age, data = Runners, alpha = 0.2, fill = "navy") %>%
  gf_freqpoly( ~ age) %>%
  gf_labs(x = "age (years)", title = "Age of runners") %>%
  gf_lims(x = c(20, 80)) %>%
  gf_theme(theme = theme_light)
```

## Overlapping cases

Sometimes you have so many points in a scatter plot that they obscure one another. 
The `ggplot` system provides two easy ways to deal with this: translucency and jittering.

Use `alpha:0.5` to make the points semi-translucent. If there are many points overlapping at one point, a much smaller value of alpha, say `alpha:0.01`.

Using `gf_jitter()` in place of `gf_point()` will move the plotted points to reduce overlap.  You can use both of these techniques, e.g. 

```{r}
gf_jitter(age ~ sex + alpha:0.05, data = Runners)
```


## For ggplot2 users: gf and gg together

The `gf_` functions generate a ggplot object as well as a character string
containing the `ggplot()` command to generate the graphic. This can be useful
when you want to use the `gf_` functions to remind you about how `ggplot()`
works, but you want to have the `ggplot()` commands directly in your document
for future modification.

Use `verbose = TRUE` to see the string being generated.

```{r}
gf_jitter(age ~ sex + alpha:0.05, data = Runners, verbose = TRUE)
```

