---
title: "ggformula: Another Option for Teaching Graphics in R to Beginners"
author: "Randall Pruim"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Another Option for Teaching Graphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
span.boxed {
  border:5px solid gray;
  border-radius:10px;
  padding: 5px;
}
span.invboxed {
  border:5px solid gray;
  padding: 5px;
  border-radius:10px;
  color: white;
}
table, td, th
{
border:0px;
}
</style>

```{r setup, include = FALSE}
library(dplyr)
library(lattice)
library(ggformula)
library(mosaic)
theme_set(theme_bw())
knitr::opts_chunk$set(
  fig.show = "hold",
  fig.width = 2.5,
  fig.height = 2.2
)
```

David Robinson's excellent blog entry "Teach the tidyverse to beginners"
(http://varianceexplained.org/r/teach-tidyverse) argues that a `tidyverse`
approach is the better of two prevalent options for teaching beginners:

1) "Base R first": teach syntax such as `$` and `[[ ]]`, use built-in 
functions like `ave()` and `tapply()`, and use base graphics.

2) "Tidyverse first": start from scratch with pipes (`%>%`), leverage `dplyr`
for data transformations and summarization, and use `ggplot2` for graphics.

As his title suggests, David prefers the second option and makes a strong case for it.
Indeed, if those were the only two options, I'd probably choose `tidyverse` as well.

## A third way

But these are not the only options.
Nick Horton's "Options for teaching R to beginners: a false
dichotomy?"(http://sas-and-r.blogspot.com/2017/07/options-for-teaching-r-to-beginners.html)
describes a third alternative that also "get[s] students doing powerful things quickly",
as David desires. 
This third way is based on a formula interface provided by the combination of 

 * the `lattice` package for graphics, 
 * several functions from the `stats` package for modeling 
 (e.g., `lm()`, `t.test()`), and 
 * the `mosaic` package for numerical summaries and for smoothing
over edge cases and inconsistencies in the other two components.  

Important in this approach is the syntactic similarity that the following
"formula template" brings to all of these operations.

&nbsp;

<center>
<h2><strong><span class="boxed">goal</span> ( <span class="boxed">y</span> ~ <span class="boxed">x</span> , data = <span class="boxed">mydata</span>, ... )</strong> 
</h2></center>

&nbsp;

Many important data analysis operations can be executed by filling in the 
four boxes with the appropriate information for the desired task.
This allows students to become fluent quickly with a powerful, coherant toolkit for 
data analysis.  

Nick's post illustrates this with an investigation of how the price of diamonds
depends on (among other things) color.  The similarity among the commands to
compute mean price for each of two colors, to create side-by-side boxplots, to
run a two-sample t test, and to fit a linear model are what make this approach
so compelling.

```{r results = "hide", fig.keep = "none"}
# diamonds2 was called `recoded' in Nick's post
library(dplyr)
diamonds2 <- diamonds %>% 
  filter(color == "D" | color == "J") %>%
  droplevels()

  mean(price ~ color, data = diamonds2)  
bwplot(price ~ color, data = diamonds2)
t.test(price ~ color, data = diamonds2)
    lm(price ~ color, data = diamonds2)
```

This "Less Volume, More Creativity" approach
is outlined in more detail in a recent [*R Journal* article](https://journal.r-project.org/archive/2017/RJ-2017-024/)
and has worked well for a growing number of instructors in first
(and subsequent) courses 
(see for example Wang et al., 
"Data Viz on Day One: Bringing Big Ideas into Intro Stats Early and Often" (2017),
TISE).

## Trouble in paradise

But as Nick hinted in his blog post, the use of `lattice` has some drawbacks.
While basic graphs like histograms, boxplots, scatterplots, and 
quantile-quantile plots are simple to make with `lattice`, it is challenging 
to combine these simple plots into more complex plots or to plot data from multiple
data sources.  Splitting data into subgroups and either overlaying with multiple 
colors or separating into sub-plots (facets) is easy, but the labeling of 
such plots is not as convenient (and takes more space) that the equivalent plots
made with `ggplot2`.  And in our experience, students generally find the look of 
`ggplot2` graphics more appealing.

On the other hand, introducing `ggplot2` into a first course is challenging. 
The syntax is more verbose, so it takes up more of the limited space on projected 
images and course handouts.  More importantly, the syntax is entirely unrelated 
to the syntax used for other aspects of the course. 
For those adopting a "Less Volume, More Creativity" approach,
`ggplot2` is tough to justify.


## ggformula: The third-and-a half way

Danny Kaplan and I recently introduced `ggformula`, a package that provides a 
formula interface to `ggplot2` graphics.  Our hope is that this provides 
the best aspects of `lattice` (the formula interface and lighter syntax) 
and `ggplot2` (modularity, layering, and better visual aesthetics).  

For simple plots, the only thing that changes is the name of the plotting function.
Each of these functions begins with `gf`.  Here are two examples, either of which 
could replace the side-by-side boxplots made with `lattice` in Nick's post.
```{r}
gf_boxplot(price ~ color, data = diamonds2)
 gf_violin(price ~ color, data = diamonds2)
```

If we like, we can even overlay these two types of plots to see how they compare.  To
do so, we simply place the then operator (`%>%`, also called a pipe) between the
two layers and adjust the transparency so we can see both where they overlap.

```{r}
gf_boxplot(price ~ color, data = diamonds2, alpha = 0.05) %>%
 gf_violin(price ~ color, data = diamonds2, alpha = 0.3, fill = "navy", color = NA)
```


## Comparing groups

Groups can be compared either by overlaying multiple groups distinguisable 
by some attribute (e.g., color) 

```{r}
gf_point(price ~ carat, color = ~ color, data = diamonds2, alpha = 0.2) %>%
  gf_lm()
```

or by creating multiple plots arranged in 
a grid rather than to overlay subgroups in the same space. 
`ggformula` provides two ways to create these facets. 
The first uses `|` very much like `lattice` does.

```{r fig.width = 7}
gf_point(price ~ carat | color, data = diamonds2, alpha = 0.2) %>%
  gf_lm()
gf_point(price ~ carat | color ~ clarity, data = diamonds2, alpha = 0.2) %>%
  gf_lm()
```

Notice that the `gf_lm()` layer inherits information from the the `gf_points()` 
layer in these plots, saving some typing when the information is the same in
multiple layers.

(The second way to adds facets with `gf_facet_wrap()` or `gf_facet_grid()`
and can be more convenient for complex plots or when customization
of facets is desired.)
      
## Fitting into the tidyverse work flow

`ggformala` also fits into a tidyverse-style workflow (arguably better than
`ggplot2` itself does).
Data can be piped into the initial call to a `ggformula` function and there 
is no need to switch between `%>%` and `+` when moving from data transformations
to plot operations.

```{r, fig.height = 1.5}
diamonds %>%
  filter(color %in% c("D", "G", "J")) %>%
  droplevels() %>%
  gf_bar( ~ color)
```

## Summary

The "Less Volume, More Creativity" appraoch based on a common formula template
has served well for several years, but the arrival of `ggformula` strengthens this
approach by bringing a richer graphical system into reach for beginners without 
introducing new syntactical structures.  The full range of `ggplot2` features 
and customizations remains available, and package vignettes and tutorials describe
these in more detail.


