---
title: "Project MOSAIC migrates to ggformula"
author: "R Pruim"
date: "7/9/2018"
output:
  html_document: 
    self_contained: no
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.show = "hold",
  fig.width = 4,
  fig.height = 3,
  fig.align = "center")
library(mosaic)
theme_set(theme_bw(base_size = 10))
```

Last summer, 
Project MOSAIC announced `ggformula`, a new package that provides a formula interface to
`ggplot2` graphics in R.
(See, for example, 
[*ggformula: another option for teaching graphics in R to beginners*](https://www.r-bloggers.com/ggformula-another-option-for-teaching-graphics-in-r-to-beginners/).)
This package provides a happy medium between `lattice` and `ggplot2` that allows
beginners to "do powerful things quickly" by adopting the formula interface of
`lattice` and R's statistical modeling functions as a means to produce `ggplot2`
graphics.


Over the past year, our experience with `ggformula` in our classes and in
faculty development workshops together with the feedback we have received from
other users have demonstrated `ggformula` to be flexible, yet easy to learn. As
part of an ecosystem that emphasizes a formula interface of `lattice` and the
core R statistical modeling functions early on and adds `tidyverse` concepts
later, `ggformula` fits better with the rest of our toolkit than do either
`lattice` or `ggplot2`, providing opportunities for more creativity with less
volume.

The recent releases of several Project MOSAIC R packages (`mosaic`,
`mosaicData`, `mosaicCore`, and `ggformula`) and the related `fastR2` package
mark the official migration of Project MOSAIC from `lattice` to `ggformula` as
its primary graphics system.  Future development includes plans to release
an updated version of `mosaicModel` which will interoperate with `ggformula`
and a new package called `ggformulaExtra` (currently only available via Github)
which adds additional functionality but relies on additional packages beyond `ggplot2`.


Many of the recent changes to the Project MOSAIC suite of packages will go largely
unnoticed by most users but were necesary to allow `ggformula` to interoperate
with the newest version of `ggplot2`. Among the small number of more noticeable
changes are a change in `gf_smooth()` so that it no longer displays confidence
bands by default (use `se = TRUE` to turn them on), expanded support for "rugs",
and the addition of `gf_sf()` for improved support for choropleth maps (based on
the new `geom_sf()` in `ggplot2`). Along the way, we also did some light
housekeeping (improving documentation, etc.) and migrated most of our package
examples from `lattice` to `ggformula`.

The basic form of the formula interface is 

```{r, eval = FALSE}
goal(y ~ x, data = myData)
```

which corresponds to SAS code like

```
PROC GOAL DATA = MYDATA; MODEL Y = X; RUN;
```

`goal()` can be replaced by a graphing (e.g., `gf_point()`) or modeling
(e.g., `lm()`) function with the number of variables involved in the formula
varying with the complexity of the plot or model desired.

```{r} 
library(mosaic)              # load the mosaic package (and ggformula)
gf_point(length ~ width, data = KidsFeet)                  # scatter plot 
      lm(length ~ width, data = KidsFeet) %>% msummary()   # linear model
```


Users of `lattice`-based Project MOSAIC materials should have little
trouble migrating to `ggformula` since the types of plots that were easiest to
construct with `lattice` can be created very similarly using `ggformula`.  For
example, the following two commands are essentially equivalent (although the
resulting plots have a different appearence).

```{r, histograms}
    histogram( ~ age | sex, data = HELPrct,    width = 2, col  = "navy")
gf_dhistogram( ~ age | sex, data = HELPrct, binwidth = 2, fill = "navy")
```

It is much simpler, however, to create complex plots using `ggformula` because
multiple layers can be stacked using the `maggrittr` pipe (`%>%`, which we
often read as "then") familiar to users of the `tidyverse` suite of packages (and
many others as well).

```{r, multilayer}
gf_jitter(Sepal.Length ~ Sepal.Width, data = iris, color = ~ Species) %>%
  gf_density2d(alpha = 0.4) %>%
  gf_jitter(geom = "rug", alpha = 0.7) %>%
  gf_lm(linetype = "dashed") %>%
  gf_refine(scale_color_brewer(type = "qual"))
```

As part of the migration to `ggformula`, a number of [related resources](https://projectmosaic.github.io/mosaic/articles/mosaic-resources.html)
have been or are being converted from `lattice` to `ggformula` as well. These
include [companion volumes for several popular statistics text books](https://projectmosaic.github.io/mosaic/articles/mosaic-resources.html#textbook-related),
our series of ["Little Books"](https://projectmosaic.github.io/mosaic/articles/mosaic-resources.html#project-mosaic-little-books), 
the [*Minimal R Vignette*](https://cran.r-project.org/web/packages/mosaic/vignettes/MinimalRgg.pdf),
and a 
[side-by-side comparison of `lattice` and `ggformula`](https://projectmosaic.github.io/mosaic/articles/ggformula-lattice.html).
In addition, the second
edition of [*Foundations and Applications of Statistics*](https://bookstore.ams.org/amstext-28/)
(Pruim, 2018) uses `ggformula` throughout.

An eventual migration from `ggformula` to native `ggplot2`, while not strictly
necessary (since the same plots can be made in either system), is easier than
the migration from `lattice` since the underlying grammar and much of the
nomenclature of `ggformula` is borrowed from `ggplot2`.  In the meantime,
equivalent `ggformula` code is generally less verbose and simpler for novices to
understand and produce.  And the use of `%>%` for layering avoids the errors
that creap in when moving between `tidyverse`, which also uses `%>%`, and
`ggplot2` which uses `+`.  Indeed, data flows can be directed seamlessly into
`ggformula` plotting commands.  This can be useful as a debugging step when
creating data pipelines or as a way to create a plot for which there is no
need to save the pre-processed data.

```{r, Galton}
Galton %>%
  filter(sex == "M") %>%  # select only male adult children
  group_by(family) %>%      #
  sample_n(1) %>%           # choose only one male from each family
  ungroup %>%               #
  mutate(                     # compute z-scores for parents' heights
    zfather = round(mosaic::zscore(father), 2),
    zmother = round(mosaic::zscore(mother), 2)
  ) %>% 
  gf_jitter(zfather ~ zmother, alpha = 0.5, 
            title = "Standardized heights of parents",
            caption = "Source: Galton") %>%
  gf_lm()
```


It has been over a year since I have used either `lattice` or `ggplot2` for
anything other than comparison examples. My co-authors and I have found the
switch from `lattice` to `ggformula` to be both straightforward (for us) and
advantageous (for our students). We encourage you to give it a try in your own
work and with your students.
